---
title: "Spatial_exam"
author: "Malene"
date: "2025-05-26"
output: html_document
---

## Loading in data
```{r}
library(readxl)
```

```{r}
population <- read_excel("FOLK1AM.xlsx", skip = 2)

#12. måned 2024
colnames(population) <- c("In total", "Municipality", "Population")
```


```{r}

students <- read_excel("SKOLM02B.xlsx", skip = 2)

#2023:2024
colnames(students) <- c("In total", "Municipality", "N_Students")
```


```{r}
library(tidygeocoder)
library(tidyverse)
library(dplyr)
library(osmdata)
library(sf)

geocoded_schools <- readRDS("music_schools_geocoded.rds")

```


## Loading the shape of Denmark and its municipalities

```{r}
library(raster)
library(geodata)

dk_admin <- gadm(country = "DNK", path = ".", level = 2) # Level 2 = municipalities

dk_admin$NAME_2[31] <- "Aarhus"
dk_admin$NAME_2[60] <- "Vesthimmerlands"
```



## Visualizing the schools and municipalities

```{r}
dk_admin <- sf::st_as_sf(dk_admin)

# Convert geocoded schools to spatial object (sf)
schools_4326 <- st_as_sf(geocoded_schools, coords = c("longitude", "latitude"), crs = 4326) #geocoded_sf

```


```{r}
library(ggplot2)
library(ggspatial)

ggplot() +
  geom_sf(data = dk_admin, fill = "white", color = "gray40") +
  geom_sf(data = schools_4326, color = "royalblue1", size = 1.5, alpha = 0.7) +
  theme_minimal() +
  annotation_scale(location = "br", width_hint = 0.25, text_cex = 0.8) +   # Scale bar
  annotation_north_arrow(location = "tr", which_north = "true",
                         style = north_arrow_fancy_orienteering) +        # Compass
  labs(title = "Music Schools in Denmark", subtitle = "Overlaid on Municipal Boundaries")
```


## Plotting number of students and music schools

```{r}
students$Municipality[14] <- "Høje Taastrup"
population$Municipality[13] <- "Høje Taastrup"

#Joining dataframes students and dk_admin
dk_students <- dk_admin %>%
  left_join(students, by = c("NAME_2" = "Municipality"))

dk_students_pop <- dk_students %>% 
  left_join(population, by = c("NAME_2" = "Municipality"))
```


```{r}
ggplot() +
  geom_sf(data = dk_students, aes(fill = N_Students), color = "white") +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90", name = "Number of Students") +
  geom_sf(data = schools_4326, color = "black", size = 1.5, alpha = 0.4) +
  theme_minimal() +
  labs(title = "Number of Students per Municipality in Denmark",
       subtitle = "With Music School Locations",
       fill = "Students")
```


### Interactive Mapping

```{r}
library(leaflet)

leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = dk_students,
              fillColor = ~colorNumeric("plasma", N_Students)(N_Students),
              color = "#444444", weight = 1, smoothFactor = 0.5,
              opacity = 1.0, fillOpacity = 0.7,
              popup = ~paste0(NAME_2, ": ", N_Students, " students")) %>%
  addCircleMarkers(data = schools_4326,
                   radius = 5, color = "firebrick",
                   label = ~name,
                   popup = ~paste0(name, "<br>", address))
```


### Relative number of students and municipalities

```{r}
dk_students_pop <- dk_students_pop %>%
  mutate(students_per_1000 = (N_Students / Population) * 1000)


ggplot() +
  geom_sf(data = dk_students_pop, aes(fill = students_per_1000), color = "white") +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90", name = "Number of Students") +
  geom_sf(data = schools_4326, color = "black", size = 1, alpha = 0.5) +
  theme_minimal() +
  annotation_scale(location = "br", width_hint = 0.25, text_cex = 0.8) +   # Scale bar
  annotation_north_arrow(location = "tr", which_north = "true",
                         style = north_arrow_fancy_orienteering) +        # Compass
  labs(title = "Number of Students per 1000 Inhabitants for Municipalities in Denmark",
       subtitle = "With Music School Locations",
       fill = "Students")
```


```{r}
library(dplyr)

# Join population data to spatial object
municipalities_joined <- dk_admin %>%
  left_join(population, by = c("NAME_2" = "Municipality"))

# Then plot a choropleth map:
ggplot(municipalities_joined) +
  geom_sf(aes(fill = Population)) +
  scale_fill_viridis_c() +
  labs(title = "Population Distribution by Municipality",
       fill = "Population") +
  theme_minimal()
```




## Newspapers research

```{r}
library(tidyverse)

#Uncomment if you need to save a new dataset:

#newspapers <- read_csv("https://labs.statsbiblioteket.dk/labsapi/api/aviser/export/fields?query=musikskole%20AND%20py%3A%5B1800%20TO%201900%5D&fields=link&fields=recordID&fields=timestamp&fields=pwa&fields=cer&fields=fulltext_org&fields=pageUUID&fields=editionUUID&fields=titleUUID&fields=editionId&fields=familyId&fields=newspaper_page&fields=newspaper_edition&fields=lplace&fields=location_name&fields=location_coordinates&max=1600&structure=comments&structure=header&structure=content&format=CSV", skip = 6)

#saveRDS(newspapers, "newspapers.rds")

newspapers <- readRDS("newspapers.rds")

```


```{r}
library(tidyverse)
library(sf)

#Split locations into multiple points
newspapers_4326 <- newspapers %>%
  mutate(
    coord_list = str_split(location_coordinates, "\\\\n")
  ) %>%
  unnest(coord_list) %>%
  separate(coord_list, into = c("lon", "lat"), sep = ",", convert = TRUE) %>%
  filter(!is.na(lon) & !is.na(lat)) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

newspapers_4326$year <- lubridate::year(newspapers_4326$timestamp)

newspapers_4326 <- newspapers_4326 %>%
  filter(str_detect(tolower(fulltext_org), "musikskole"))

```


```{r}

ggplot() +
  geom_sf(data = dk_admin, fill = "white", color = "gray70") +
  geom_sf(data = newspapers_4326, aes(color = year), alpha = 0.3, size = 1.5) +
  scale_color_viridis_c(option = "plasma") +
  theme_minimal() +
  annotation_scale(location = "br", width_hint = 0.25, text_cex = 0.8) +   # Scale bar
  annotation_north_arrow(location = "tr", which_north = "true",
                         style = north_arrow_fancy_orienteering) +        # Compass
  labs(title = "Mentions of 'Musikskole' in Historical Newspapers",
       subtitle = "Geocoded mentions by year",
       color = "Year")
```


```{r}
newspapers_4326 %>%
  count(year) %>%
  ggplot(aes(x = year, y = n)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Frequency of 'Musikskole' Mentions Over Time",
       x = "Year", y = "Number of Mentions")
```


```{r}
library(ggspatial)

ggplot() +
  geom_sf(data = dk_admin, fill = "white", color = "black") +
  stat_density_2d(data = st_coordinates(newspapers_4326) %>% as.data.frame(),
                  aes(X, Y, fill = after_stat(level)),
                  geom = "polygon", alpha = 0.5) +
  scale_fill_viridis_c() +
  coord_sf() +
  labs(title = "Density of 'Musikskole' Mentions") +
  theme_minimal() +
  annotation_scale(location = "br", width_hint = 0.25, text_cex = 0.8) +   # Scale bar
  annotation_north_arrow(location = "tr", which_north = "true",
                         style = north_arrow_fancy_orienteering)        # Compass
  
```


```{r}
# Add a decade column
newspapers_4326$decade <- floor(newspapers_4326$year / 10) * 10

newspapers_25832 <- st_transform(newspapers_4326, 25832)  # UTM zone 32N for Denmark # newspapers_transformed

ggplot() +
  geom_sf(data = dk_admin, fill = "white", color = "black") +
  stat_density_2d(data = st_coordinates(newspapers_4326) %>% as.data.frame(),
                  aes(X, Y, fill = after_stat(level)),
                  geom = "polygon", alpha = 0.5) +
  geom_sf(data = schools_4326, color = "black", size = 2) +
  scale_fill_viridis_c() +
  coord_sf() +
  labs(title = "Density of 'Musikskole' Mentions") +
  theme_minimal() +
  annotation_scale(location = "br", width_hint = 0.25, text_cex = 0.8) +   # Scale bar
  annotation_north_arrow(location = "tr", which_north = "true",
                         style = north_arrow_fancy_orienteering)        # Compass
  

```


## Temporal map

```{r}
library(gganimate)

# Reproject to appropriate CRS for spatial density (e.g., UTM)
newspapers_transformed <- st_transform(newspapers_4326, 25832)
dk_admin_transformed <- st_transform(dk_admin, 25832)

# Convert coordinates to data frame with year
density_data <- st_coordinates(newspapers_transformed) %>%
  as.data.frame() %>%
  mutate(year = newspapers_transformed$year)

# Plot with animation by year
p <- ggplot() +
  geom_sf(data = dk_admin_transformed, fill = "white", color = "black") +
  stat_density_2d(
    data = density_data,
    aes(X, Y, fill = after_stat(level)),
    geom = "polygon", alpha = 0.4, contour = TRUE
  ) +
  scale_fill_viridis_c() +
  coord_sf() +
  labs(title = "Density of 'Musikskole' Mentions", subtitle = "Year: {closest_state}") +
  theme_minimal() +
  transition_states(year, transition_length = 3, state_length = 5) +
  ease_aes('linear')

gganimate::animate(p, width = 800, height = 600, fps = 5)

anim_save("music_school_density.gif")
```


### Interactive Map to see mentions of 'musikskole' (black) per decade vs modern music schools (orange)


```{r}
# Libraries
library(leaflet)
library(dplyr)
library(sf)
library(lubridate)

# Base leaflet map
map <- leaflet() %>%
  addTiles()

# Add schools layer
map <- map %>%
  addCircleMarkers(data = schools_4326,
                   radius = 2,
                   color = "black",
                   fillOpacity = 0.7,
                   popup = ~name)  

# Add newspaper markers by decade
for (decade in sort(unique(newspapers_4326$decade))) {
  decade_data <- newspapers_4326 %>% filter(decade == !!decade)

  map <- map %>%
    addCircleMarkers(
      data = decade_data,
      radius = 2,
      color = "orange",
      fillOpacity = 0.2,
      group = paste0("Newspapers ", decade),
      popup = ~paste("Year:", year)
    )
}

# Add layer control to toggle decades
map <- map %>%
  addLayersControl(
    overlayGroups = paste0("Newspapers ", sort(unique(newspapers_4326$decade))),
    options = layersControlOptions(collapsed = FALSE)
  )

# Display map
map

```




```{r}

# Spatial join to municipalities
newspapers_joined <- st_join(newspapers_4326, dk_admin)

mention_counts <- newspapers_joined %>%
  st_drop_geometry() %>%
  count(NAME_2)

# Join count data back to geometry for plotting
kommuner_count <- dk_admin %>%
  left_join(mention_counts, by = "NAME_2")

# Plot
ggplot(kommuner_count) +
  geom_sf(aes(fill = n), color = "grey60") +
  scale_fill_viridis_c(option = "inferno", na.value = "grey80") +
  theme_minimal() +
  labs(
    title = "Mentions of 'Musikskole' per Municipality",
    fill = "Mentions"
  )

```


## Statistical Analysis

### Proximity Analysis

```{r}
schools_transformed <- st_transform(schools_4326, crs = 25832) #geocoded_sf_transformed

# Calculate nearest school for each newspaper mention
nearest_index <- st_nearest_feature(newspapers_transformed, schools_transformed)

# Get distance (in meters)
distances <- st_distance(newspapers_transformed, schools_transformed[nearest_index, ], by_element = TRUE)

# Add distance as a column
newspapers_transformed$distance_to_school <- as.numeric(distances)

```


```{r}
library(ggplot2)

ggplot(newspapers_transformed, aes(x = distance_to_school)) +
  geom_histogram(bins = 30, fill = "royalblue1", color = "white") +
  labs(title = "Distance from Newspaper Mentions to Nearest Music School",
       x = "Distance (m)", y = "Count") +
  theme_minimal()

```



### Spatial Autocorrelation

```{r}
library(spdep)

# Create spatial weights
coords_sp <- as_Spatial(newspapers_transformed)
nb <- knn2nb(knearneigh(coords_sp, k = 5))
lw <- nb2listw(nb, style = "W")

# Global Moran's I
moran.test(newspapers_transformed$distance_to_school, lw)

```


### Monte Carlo - Are the newspapers closer to the music schools than by chance?

```{r}
dist_mentions <- st_distance(newspapers_transformed, schools_transformed)
min_dist_mentions <- apply(dist_mentions, 1, min)  # shortest distance per mention
```

```{r}
set.seed(8)
n_random <- nrow(newspapers_transformed)  # same number as mentions

random_points <- st_sample(dk_admin_transformed, size = n_random, type = "random")
random_points <- st_as_sf(random_points)
```

```{r}
dist_random <- st_distance(random_points, schools_transformed)
min_dist_random <- apply(dist_random, 1, min)
```

```{r}
library(ggplot2)

df <- data.frame(
  distance = c(as.numeric(min_dist_mentions), as.numeric(min_dist_random)),
  group = rep(c("Mentions", "Random"), each = n_random)
)

ggplot(df, aes(x = distance / 1000, fill = group)) +  # convert to km
  geom_density(alpha = 0.5) +
  labs(title = "Distance to Nearest Music School",
       x = "Distance (km)", y = "Density") +
  theme_minimal()

```

```{r}
wilcox.test(min_dist_mentions, min_dist_random)
```

The p-value < 0.05, which suggests that mentions of 'musikskole' in danish newspapers are closer to the music schools than by chance. 


```{r}
set.seed(8)
n_sim <- 1000
n_random <- nrow(newspapers_transformed)

median_dists <- numeric(n_sim)

for (i in 1:n_sim) {
  random_points <- st_sample(dk_admin_transformed, size = n_random, type = "random") %>% st_as_sf()
  dist_random <- st_distance(random_points, schools_transformed)
  min_dists <- apply(dist_random, 1, min)
  median_dists[i] <- median(min_dists)
}

# Now compare:
real_median <- median(min_dist_mentions)

# Plot
ggplot(data.frame(median_dists = median_dists), aes(x = median_dists / 1000)) +
  geom_histogram(bins = 30, fill = "royalblue1", color = "white") +
  geom_vline(xintercept = real_median / 1000, color = "red", linetype = "dashed") +
  labs(title = "Monte Carlo Simulation of Median Distances",
       x = "Median Distance to Nearest Music School (km)", y = "Frequency") +
  theme_minimal()

```





## Hotspot Analysis

### Hotspots faceted by decade
```{r}

ggplot() +
  geom_sf(data = dk_admin, fill = "white") +
  stat_density_2d(data = st_coordinates(newspapers_4326) %>%
                    as.data.frame() %>%
                    bind_cols(decade = newspapers_4326$decade),
                  aes(X, Y, fill = after_stat(level)),
                  geom = "polygon", alpha = 0.5) +
  scale_fill_viridis_c() +
  facet_wrap(~ decade) +
  coord_sf() +
  labs(title = "Hotspots of 'Musikskole' Mentions by Decade") +
  theme_minimal()

```

### Gi*

```{r}
library(sf)
library(spdep)
library(tidyverse)
library(tmap)

# Create a grid
grid <- st_make_grid(
  newspapers_transformed,
  cellsize = 8000,  # 8 km
  square = FALSE     # hexagonal grid
) %>% st_sf()

# Add ID
grid$grid_id <- 1:nrow(grid)

# Spatial join to count mentions per grid
grid_mentions <- st_join(grid, newspapers_transformed, join = st_contains)

# Count number of mentions per grid cell
mention_counts <- grid_mentions %>%
  group_by(grid_id) %>%
  summarise(count = n())

# Merge count back with original grid
grid <- st_join(grid, mention_counts, by = "grid_id")
grid$count[is.na(grid$count)] <- 0

# Create neighbor list from polygons
nb <- poly2nb(grid)  

# Convert to spatial weights
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# Run Gi* statistic on mention count
gi_star <- localG(grid$count, lw, zero.policy = TRUE)

# Add the result to the grid
grid$gi_star <- as.numeric(gi_star)

library(tmap)

tmap_mode("view")

tm_shape(grid) +
  tm_polygons("gi_star",
              palette = "-RdBu",
              style = "quantile",
              title = "Gi* Hotspot Score") +
  tm_layout(title = "Hotspot Analysis of 'musikskole' Mentions")


```


```{r}
# Add grid_id
grid$grid_id <- 1:nrow(grid)

# Convert schools to sf
#schools_sf <- st_as_sf(geocoded_schools, coords = c("longitude", "latitude"), crs = 4326)
#schools_sf <- st_transform(schools_sf, st_crs(grid))

# Assign each school to a grid cell
grid_schools <- st_join(schools_transformed, grid, join = st_within)

# Count schools per grid cell
school_counts <- grid_schools %>%
  st_drop_geometry() %>%
  group_by(grid_id) %>%
  summarise(schools_count = n())

# Merge counts back to grid
grid <- left_join(grid, school_counts, by = "grid_id")
grid$schools_count[is.na(grid$schools_count)] <- 0

```


```{r}
cor.test(grid$count, grid$schools_count, method = "spearman")
```

```{r}
# Use spatial weights from earlier
moran.test(grid$schools_count, listw = lw)
```





